\section{Introduction}
The Ethereum~\cite{WhitePaper} platform is a system that appears as a
singleton networked computer usable by anyone, but is actually built as a
distributed database that utilizes blockchain technology to achieve consensus.
%
One of the features that sets Ethereum apart from other blockchain systems is
the ability to store and execute code inside this database, via the Ethereum
Virtual Machine (\emph{EVM}).
%
In contrast to traditional server systems, anyone can inspect this stored code
and execute functions that can have stateful effects.
%
Since blockchains are typically used to store ownership relations of valuable goods
(for example
cryptocurrencies),
malicious actors have a monetary incentive to analyze the inner
workings of such code. Because of that, testing (i.e.\  dynamic analysis of some
typical inputs) does not suffice and analyzing all possible inputs by utilizing
static analysis or formal verification is recommended.

SAT/SMT-based techniques have been used extensively for program 
verification~\cite{Biere99,Donaldson11,Komuravelli13,Beyer11,Kroening14,Alt17}.
%
This paper shows how the Solidity compiler, which generates EVM bytecode,
utilizes an SMT solver and a Bounded Model Checking~\cite{Biere99} (BMC) approach to
verify safety properties that can be specified as part of the source code, as
well as fixed targets such as arithmetic underflow/overflow, division by zero and detection of
unreachable code and trivial conditions.
%
For the user, the main advantage of this system over others is that they do not
need to learn a second verification language or how to use any new tools, since
verification is part of the compilation process.
%
The Solidity language has requirement and assertion constructs that enable
assuming and checking conditions at run-time.  The verification component
builds on top of this and tries to
verify at compile-time that the asserted conditions hold for any input,
assuming the given requirements.

This paper is organized as follows:  Sec.~\ref{section:smart_contracts}
introduces the EVM and smart contracts.  Sec.~\ref{section:solidity} gives a
very brief overview of Solidity. Sec.~\ref{section:smt} discusses the
translation from Solidity to SMT statements and next challenges. Finally,
Sec.~\ref{section:conclusion} contains our concluding remarks.

\begin{paragraph}{Related work.}
Oyente~\cite{Luu2016}, Mythril~\cite{Mythril} and MAIAN~\cite{NKSSH18} are SMT-based symbolic
execution tools for EVM bytecode that check for specific known vulnerabilities,
where Oyente also checks for assertion fails. They simulate the virtual
machine and execute all possible paths, whereas this work follows a BMC approach.

Subsets of Solidity have been translated to Why3~\cite{Why3},
F*~\cite{Bhargavan2016} and LLVM~\cite{ZEUS}, with the goal of verifying
statements in those languages instead, where the first requires
learning a new annotation specification language; the second only verifies
fixed vulnerability patterns and does not verify custom user-provided
assertions; and the latter focuses on discovering non-ideal coding patterns
instead of actual vulnerabilities that demand a fix.

\end{paragraph}
