\section{SMT-based Solidity verification}
\label{section:smt}

SMT solvers are powerful tools to prove satisfiability of formulas in different
logics which often have the necessary expressiveness to model software in a
straightforward manner~\cite{Komuravelli13,Alt17,Donaldson11,Beyer11}.

We translate Solidity contracts and their functions into SMT formulas using a
combination of different quantifier-free theories.
%
We shall name the translated formulas the \emph{SMT encoding} of the Solidity
program.
%
The goal of the translation from Solidity to SMT formulas is to verify safety
properties from the Solidity program by performing queries to the SMT solver.

\subsection{SMT encoding}

The SMT encoding is computed during a depth-first traversal of the
abstract syntax tree (AST) of the
Solidity program and thus roughly follows the execution order.
%
For now, each function is analyzed in isolation and thus the 
context regarding the SMT solver (contract storage, local variables, etc.)
is cleared before each function of a contract is visited.
%
There are five types of formulas that are encoded from Solidity inside each function.
%
Three of them, \emph{Control-flow}, \emph{Type constraint} and \emph{Variable assignment}
are simply translated as SMT constraints.
%
The \emph{Branch conditions} are the conditions of the current branch of
execution and thus grow and shrink as we traverse the AST.
%
The last, \emph{Verification Target}, creates a formula consisting of the
verification goal conjoined with the previously mentioned constraints,
including the current branch conditions, and
queries the SMT solver for satisfiability.
%
The different types of encoding are described below.

\begin{paragraph}{Branch conditions.}
For an if-statement \code{if ($c$) $T$ else $F$}, we add
$c$ to the branch conditions during the visit of $T$.
After that, we replace $c$ by $\neg c$
for the visit of $F$ and also remove that when we are
finished with the if-statement.
\end{paragraph}
  
\begin{paragraph}{Control-flow.}
These constraints model conditional termination of execution.
A \code{require($r$)} statement (and similar for \code{assert($r$)})
terminates execution if $r$ evaluates to $\mathrm{false}$, but
of course only if it is executed.
%
Let $b$ be the conjunction of the current branch conditions. We separately check
satisfiability of the formulas $b \land r$ for \code{require} 
(unreachable code if unsatisfiable) and
$b \land \neg r$ for \code{assert} (assertion failure if unsatisfiable),
and add $b \to r$ to the set of constraints. Note that due to the implication,
we can keep this constraint even when we leave the current branch.
\end{paragraph}

\begin{paragraph}{Type constraint.}
A variable declaration leads to a correspondent SMT variable that is assigned
the default value of the declared type.
%
For example, Boolean variables are assigned $\mathrm{false}$, and integer variables are
assigned 0.
%
Function parameters are initialized with a range of valid values for the given
type, since their value is unknown.  For instance, a parameter \code{uint32 x}
is initialized as $0 \le x < 2^{32}$ (32 bits), a parameter \code{int256 y}
is initialized as $-2^{255} \le y < 2^{255}$, and a parameter \code{address
a} is assigned the range $0 \le a < 2^{(8*20)}$ (20 bytes).
%
The encoder currently supports Boolean and the various sizes of Integer
variables.

\end{paragraph}

\begin{paragraph}{Variable assignment.}
The encoding of a variable assignment follows the \emph{Single Static Assignment}
(SSA) where each assignment to a program variable introduces a new SMT variable
that is assigned only once.
%
When a program variable is modified inside different branches of execution,
a new variable is created after the branch to re-combine the different
values after the branches.
We use the if-then-else-function \code{ite} to assign the value
\code{ite(c, $x_1$, $x_2$)} (if-then-else), where $c$ is the branch
condition and $x_1$ and $x_2$ are the two SSA variables corresponding to $x$
at the ends of the branches (cf.\ the $\phi$ function in SSA).
\end{paragraph}

\begin{paragraph}{Verification target.}
Every arithmetic operation is checked against underflow and overflow according
to the type of the values, and an example is given if there is an underflow or
overflow.
%
We also check whether branch conditions are constant, warning the user about
unreachable blocks or trivial conditions.
%
The conditions in calls to \code{assert} represent target postconditions that the
Solidity programmer wants to ensure at runtime and are verified statically.
%
If it is possible to disprove the assertion provided that the control flow
can reach it (i.e.\ the current branch conditions are satisfiable),
the user is given a counterexample.
%
In contrast, \code{require} conditions are meant to be used as filters for
unwanted input values when they are unknown, for example, in public functions,
acting like preconditions for the rest of the scope.
%
Therefore, failing calls to \code{require} are not treated as errors and are
just checked for triviality and reachability.
\end{paragraph}

\begin{figure}
\noindent\begin{minipage}{.48\textwidth}
\begin{verbatim}
contract C
{
  function f(uint256 a, uint256 b)
  {
    if (a == 0)
      require(b <= 100);
    else if (a == 1)
      b = 1000;
    else
      b = 10000;
    assert(b <= 100000);
  }
}
\end{verbatim}
\end{minipage}\hfill
\begin{minipage}{.48\textwidth}
1. $a_0 \ge 0 \land a_0 < 2^{256}  \land \phantom{x}$\\
2. $b_0 \ge 0 \land b_0 < 2^{256}  \land \phantom{x}$\\
3. $(a_0 = 0) \rightarrow (b_0 \le 100) \, \land$\\
4. $b_1 = 1000 \land b_2 = 10000$\\
5. $b_3 = ite(a == 1, b_1, b_2) \land \phantom{x}$\\
6. $b_4 = ite(a == 0, b_0, b_3) \land \phantom{x}$\\
7. $\neg b_4 \le 100000$
\end{minipage}
\caption{SMT encoding of an assertion check.}
\label{figure:solidity_encoding_1}
\end{figure}

Figure~\ref{figure:solidity_encoding_1} shows on the left a Solidity sample
that requires all five types of encoding, shown on the right, in order to
verify the intended properties.
%
Since the variables \code{uint256 a} and \code{uint256 b} are function
parameters, they are initialized (lines 1 and 2) with the valid range of values
for their type (\code{uint256}).
%
If \code{a = 0}, the \code{require} condition about \code{b} is used as a
precondition when verifying the assertion in the end of the function (line 3).
%
The next two assignments to \code{b} create the new SSA variables
$b_1$ and $b_2$ (line 4).
%
Variable $b_3$ encodes the second and third conditions, and $b_4$
encodes the first condition (lines 5 and 6).
%
Finally, $b_4$ is used in the assertion check (line 7).
%
Note that the nested control-flow is implicitly encoded in the $ite$
variables $b_3$ and $b_4$.
%
We can see that the target assertion is safe within its function.

It is important to highlight that errors are irrelevant if they result in a
state change reversion (Sec.~\ref{section:smart_contracts}). The user should be warned
about checks such as overflow only if they do not result in a state reversion.
%
One popular example is the SafeMath~\cite{SafeMath} contract which
is commonly used to turn wrapping arithmetics into overflow-checked arithmetics:

\begin{verbatim}
  function add(uint256 a, uint256 b) internal pure
        returns (uint256) {
    uint256 c = a + b;
    require(c >= a);
    return c;
  }
\end{verbatim}

Although the tool detects an overflow in the computation of \code{a + b},
the overflow will result in a truncation of \code{c} in two's complement and thus
any execution that contains the overflow will revert at the \code{require}.
%
In this case the user should not be warned of the error, since no erroneous cases
exist in accepted executions.

As described above, the component performs several local checks during a single
run, therefore it is critical that the used SMT solver supports
incremental checking.
%
Moreover, we do not abstract difficult operations such as multiplication
between variables, and rather try to give precise answers when possible.
%
Therefore we combine various quantifier-free theories, such as Linear
Arithmetics, Uninterpreted Functions and Nonlinear Arithmetics. 
%
Solidity has integrated Z3~\cite{Z3} and CVC4~\cite{CVC4} via their C++ APIs.
%
The two SMT solvers are used together to increase solving power.
%
This has been important especially for the programs that require Nonlinear
reasoning, since often one solver is able to prove a property that the other
cannot.
%
The component is also able to generate \code{smtlib2}~\cite{SMTLIB}
formulas in order to interface with additional solvers.


\subsection{Future plans}

Our current implementation plans for the component involve supporting
a larger subset of the language, including more complex data structures
such as \code{mapping}.
%
We intend to build the component as a Bounded Model Checker, unrolling loops
up to a constant bound and automatically detecting bounds when possible.

One of the most interesting aspects we intend to research and support is
multi-transaction invariants.
%
The ultimate goal is to compute invariants for state variables (resident in the
contract's storage) considering any arbitrary number of calls to the contract.
%
This would enable these invariants to be used as preconditions whenever they
are accessed.
%
As an example, take contract \code{Token} from Sec.~\ref{section:solidity}.
%
We can see from the constructor that the deployer of the contract receives
10000 tokens.
%
The only way to move tokens is via the function \code{transfer}, which decreases a
certain amount of tokens from one account, if it owns enough, and increases the
same amount in another account.
%
As we can see, the number of total tokens never changes and the invariant
$$\sum_{a \in \mathrm{balances}} \mathrm{balances}[a] = 10000$$ holds in the beginning of any
function of the contract.

Function \emph{modifiers} are Solidity constructs that are used as patterns to
change the behavior of functions, and in many cases, to restrict them.
%
Commonly used modifiers are, for example, allowing only the owner of the
contract to execute the function, or executing a function if and only if the
amount of Ether sent is greater than a certain value.
%
We intend to use properties inferred from modifiers as function preconditions,
therefore improving proving power.

The idea of \emph{Effective Callback Freeness} was recently introduced by
\cite{Grossman}.
%
A smart contract $C$ is effectively callback free, if any state change caused
by an internal callback can also be caused by an execution that does not have this callback.
%
The authors show that most of the contracts deployed on Ethereum have this
property.
%
This is a powerful property, since it means that any invariant computed for a
contract's state variables still holds even after calling external contracts
with unknown behavior.
%
We intend to study how to integrate this approach to our static analysis.
